Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-01 Paper/paper.tex
adding more snr threshold plots

% Macros for text notes and comments
\newcommand{\Note}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\D}{\mathrm{d}}
\newcommand{\lambdans}{\Lambda_\mathrm{NS}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\deff}{D_\mathrm{eff}}

@@

Moving on from mass to spin parameters, we now consider the measurement of spin
angular momentum of the BH $\chibh$. The ratio of the systematic bias introduced into
and the statistical uncertainty associated with $\chibh$ measurement is shown in
Fig.~\ref{fig:TN_BHspinBias_vs_Lambda_SNR}. The presentation of information in this 
figure is identical to that in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}. A diverging colormap is used because both 
extremes of the colorbar range point to large systematic biases, while its zero (or
small) value lies in the middle. For the lowest considered SNR $\rho=30$, $\chibh$
bias is negligible, being below $50\%$ of its statistical measurement uncertainty.
For the most deformable NS structure {\it and} with a low-mass BH, the statistical
bias in $\chibh$ does become an important error. This is furthered at higher SNR, 
considering $\rho=50$ next (middle column of Fig.~\ref{fig:TN_BHspinBias_vs_Lambda_SNR}).
We find that for low-mass BHs (with masses below $4.5M_\odot$), the systematic 
measurement bias for BH spin exceeds its statistical uncertainty to become the 
dominant source of error. Not surprisingly, the systematic bias at both spin 
extremes pushes towards milder spins, since the prior excludes any more extreme
spins than $[-0.75, +0.75]$ and that pushes the bulk of the posterior inwards.
However as the signal's $\lambdans$ is increased from $500\rightarrow 2000$, the 
low-mass BH-NS binaries with large spins on the BHs show very large systematic biases
in spin values, that becomes larger than the statistical uncertinty for $\rho\geq30$
and reaches up to $4\times$ the same quantity as $\rho\rightarrow 70$.

Summarizing these results, we find that irrespective of system parameters, below a
signal-to-noise ratio of $30$, our measurements of mass and spin parameters of 
astrophysical BHNS binaries will remain limited by the intrinsic uncertainty due to 
instrument noise, and not depend on whether we include tidal effects in template
models. However, when the signal-to-noise ratio exceeds $30$ the systematic bias in
binary mass and spin measurements become comparable to and exceed the uncertainty
due to noise. Of the different parameters considered, we find that the measurement 
of $\eta$ degrades worst (in a relative-error sense) amongst all other parameters
of BHNS binaries, due to the use of BH-BH waveforms in deciphering an NSBH signal.

@@

%\begin{figure*}
%\centering    
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_Lambda500_0_CI68_3.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_Lambda800_0_CI68_3.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_Lambda1000_0_CI68_3.png}\\
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_Lambda500_0_CI90_0.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_Lambda800_0_CI90_0.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_Lambda1000_0_CI90_0.png}
%\caption{This figure shows.}
%\label{fig:TT_SNRThresholdForLambdaMeasurement_BHspin_MassRatio_CI90_0}
%\end{figure*}

@@

@@

\includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}\\

@@

@@

% \begin{figure*}
% \includegraphics[trim={0 0 0 0},width=\columnwidth]{plots/TT_q2-00_mNS1-35_chiBH0-75_Lambda1500-0_SNR20-0_corner_eta_Mc_chi2_Lambda.pdf}
% \includegraphics[trim={0 0 0 0},width=\columnwidth]{plots/TT_q2-00_mNS1-35_chiBH0-75_Lambda1500-0_SNR20-0_corner_m1_m2_chi2_Lambda.pdf}
% \caption{These figure shows the $1-$ and $2-$dimensional 
% \textit{a posteriori} probability distributions 
% for different physical parameters describing a BHNS system including NS matter 
% effects, as measured by a Bayesian Markov-chain Monte Carlo using tidal templates.
% The true parameters of the injected signal are: 
% $\mathcal{M}=1.6406M_\odot, \eta=0.2222, \chibh=\chi_2=0.75$ and $\lambdans=1500$,
% with the injected SNR $\rho=20$. We show this case as typical of those with the largest
% tidal signatures imprinted on the emitted GWs (due to the system being comparable
% mass-ratio with a large-aligned spin on the BH and a fairly deformable NS), yet 
% with low SNR.
% % 
% We note a few things:
% (i) chirp mass is well constrained, with a nearly-Gaussian probability distribution;
% while (ii) the other $3$ less constrained parameters rail against
% prior boundaries. The specific boudary conditions that we run into are (a) for $\eta$,
% we are limited by the requirement that $m_\mathrm{NS}\geq 1.2M_\odot$ 
% of our templates; (b) for BH spin, we are limited by the $\chibh\leq+0.75$ requirement
% that results from a template model limitation~\cite{Lackey:2013axa}; and (c) for 
% $\lambdans$, we
% restrict it to $[0, 4000]$ for the same reason as (b). These prior boundaries
% limit the exploration of the entire likelihood surfaces. While (a) is astrophysically
% justified, given that we have not observed NSs with masses below 
% $1.25M_\odot$, (b) and (c) will be removed in future studies as more evolved template
% models become available. See, also, Fig.~\ref{fig:TT_logLikeProb_q2_chi75_L1500_SNR20}.
% % 
% Finally, we note that the left panel corresponds to the actual sampling parameters,
% while the right panel to derived ones.
% \label{fig:TT_pdf_q2_chi75_L1500_SNR20}
% \end{figure*}
% %
% \begin{figure*}
% % \includegraphics[width=.66\columnwidth]{plots/TT_q2-00_mNS1-35_chiBH0-75_Lambda1500-0_SNR20-0_corner_eta_Mc_chi2_Lambda.pdf}
% \includegraphics[width=\columnwidth]{plots/TT_logLike_vs_m1_m2_eta0-22_Mc1-64_chi0-00_chi0-50_L1000-0_SNR30.pdf}
% \includegraphics[width=\columnwidth]{plots/TT_logProb_vs_m1_m2_eta0-22_Mc1-64_chi0-00_chi0-50_L1000-0_SNR30.pdf}
% \caption{These figures show the effect of limitation (a) (described in the caption
% of Fig.~\ref{fig:TT_pdf_q2_chi75_L1500_SNR20}) on the likelihood surface portion
% that is actually accessible to our Bayesian inferencing algorithm. On the left we show the
% entire likelihood surface in the $\mbh-\mns$ plane, while on the left is the 
% portion not forbidden by the requirement $\mns\geq 1.2M_\odot$. Similarly, our
% priors on BH spin and NS deformability further curtail the portion of sampled
% parameter space, albeit to much less impact than this mass cut.}
% \label{fig:TT_logLikeProb_q2_chi75_L1500_SNR20}
% \end{figure*}

%
%% 
%\begin{figure*}
%\centering    
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_LambdaThresholdForLambdaMeasurement_BHspin_BHmass_SNR30_0_CI68_3.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_LambdaThresholdForLambdaMeasurement_BHspin_BHmass_SNR50_0_CI68_3.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_LambdaThresholdForLambdaMeasurement_BHspin_BHmass_SNR90_0_CI68_3.png}\\
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_LambdaThresholdForLambdaMeasurement_BHspin_BHmass_SNR50_0_CI90_0.png}
%\includegraphics[width=.65\columnwidth]{plots-TT/TT_LambdaThresholdForLambdaMeasurement_BHspin_BHmass_SNR90_0_CI90_0.png}
%\caption{This figure shows.}
%\label{fig:TT_LambdaThresholdForLambdaMeasurement_BHspin_BHmass}
%\end{figure*}

Having shown in the previous section that we begin to care for NS tidal effects
for signals in the XXX corner of the paraemter space, with SNRs above YY, here
we investigate the effect of the improvement in the accuracy of the recovered
parameters when tidal effects are included in the templates.

\begin{enumerate}
\item Where in the parameter space can we actually make a statement about 
$\Lambda_\mathrm{NS}$ ?
\item What is the reduction in the bias of the maximum likelihood parameters
when using tidal templates?\newline
``Plot the ratio of the bias between N templates and T templates (against T
signals), as a function of SNR.''
\item What is the reduction in the uncertainty in binary mass and spin, if
any?\newline
``show how the the $90\%$ confidence intervals shrink, as a function of SNR, 
when we go from using N templates to T templates.''
\end{enumerate}

\subsection{Constraining the tidal deformability of the neutron star}\label{s2:measuring_lambda}

With the enhanced sensitivity of aLIGO, we are likely to be able to measure or
constrain deviations in the observed GW signals, especially close to binary
merger, from their expected form under the point-particle approximation.
The tidal disruption of the NS close to merger, and outside of the 
inner-most circular orbit, will drastically change the morphology of the emitted
GW signal, the measurement of which will allow us to constrain the equation 
of state (EoS) of nuclear matter.
The earlier in the orbit the NS disrupts, the sharper is the drop in the emitted
GW signal for frequencies above the disruption frequency. In this section we
study the ability of aLIGO to constrain the 
$\lambdans\sim \left(R/M\right)_\mathrm{NS}^5$ parameter, which
the combination that parametrizes the deviation in the emitted GW signal.
Past studies~\cite{}
have studied a limited sample of systems with distinguishability criterion that
are typically applicable in the very high signal-to-noise ratio cases. 
In what follows, we will use stochastic parameter estimation algorithms to show
aLIGO's capability of extracting EoS information out from GW signals with 
more realistic (likely) SNRs.

As before we focus on binaries with mass-ratios $q=\{[2,3,4,5\}$, with the NS mass
fixed at $1.35M_\odot$. We allow BH spin to take the values 
$\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000\}$. For each of these
parameter combinations, we inject a tidal-corrected waveform using the 
Lackey et al~\cite{} model , and use MCMC with ensemble sampling to construct the 
posterior probability distribution for different binary parameters with (a) BHBH
templates, and (b) NSBH templates. We will discuss the results of (b) here.

>>>>

% Macros for text notes and comments
\newcommand{\Note}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\D}{\mathrm{d}}
\newcommand{\lambdans}{\Lambda_\mathrm{NS}}
\newcommand{\dlambda}{\delta\Lambda}

\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\deff}{D_\mathrm{eff}}

@@

Moving on from mass to spin parameters, we now consider the measurement of spin
angular momentum of the BH $\chibh$. The ratio of the systematic bias introduced into
and the statistical uncertainty associated with $\chibh$ measurement is shown in
Fig.~\ref{fig:TN_BHspinBias_vs_Lambda_SNR}. The presentation of information in this 
figure is identical to that in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}. A diverging colormap is used because both 
extremes of the colorbar range point to large systematic biases, while its zero (or
small) value lies in the middle. For the lowest considered SNR $\rho=30$, $\chibh$
bias is negligible, being below $50\%$ of its statistical measurement uncertainty.
For the most deformable NS structure {\it and} with a low-mass BH, the statistical
bias in $\chibh$ does become an important error. This is furthered at higher SNR, 
considering $\rho=50$ next (middle column of Fig.~\ref{fig:TN_BHspinBias_vs_Lambda_SNR}).
We find that for low-mass BHs (with masses below $4.5M_\odot$), the systematic 
measurement bias for BH spin exceeds its statistical uncertainty to become the 
dominant source of error. Not surprisingly, the systematic bias at both spin 
extremes pushes towards milder spins, since the prior excludes any more extreme
spins than $[-0.75, +0.75]$ and that pushes the bulk of the posterior inwards.
However as the signal's $\lambdans$ is increased from $500\rightarrow 2000$,
low-mass BH-NS binaries with large spins on the BH show very large systematic biases
in spin values, that become larger than the statistical uncertainty for $\rho\geq 30$
and reach up to $4\times$ the same quantity as $\rho\rightarrow 70$.

Summarizing these results, we find that irrespective of system parameters, below a
signal-to-noise ratio of $30$, our measurements of mass and spin parameters of 
astrophysical BHNS binaries will remain limited by the intrinsic uncertainty due to 
instrument noise, and do not depend on whether we include tidal effects in template
models. However, when the signal-to-noise ratio exceeds $30$ the systematic bias in
binary mass and spin measurements become comparable to and exceed the uncertainty
due to noise. Of the different parameters considered, we find that the measurement 
of $\eta$ degrades worst (in a relative-error sense) amongst all other parameters
of BHNS binaries, due to the use of BH-BH waveforms in deciphering an NSBH signal.

@@

@@

\includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor200LambdaMeasurement_BHspin_BHmass_Lambda1000_0_CI90_0}

@@

% \includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}\\

@@

\label{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}

@@

In the previous section we showed that we begin to care for the effects of NS's
deformation due to its companion hole's tidal field, if (a) the BH is sufficiently
small, (b) BH spin is aligned and $\gtrsim +0.4$, and (c) the NS is
not very compact, with $\lambdans\gtrsim 1000$. The first two conditions
enhance tidal effects on emitted GWs by increasing the number of orbits at small
separation, where the deviation of NSBH and BHBH dynamics is the largest. The third,
i.e. (c), simply makes the NS more easily deformable. All of them reduce the onset 
frequency for NS's disruption close to merger. Therefore, systems that fulfil the 
above conditions have considerably (or, at least, the most) different GW morphology
when compared to equivalent BHBH coalescences. Ultimately, the extent of 
tidal modulations of emitted GWs depends on the equation of state of neutron star
matter, which becomes an indirect measurable in GW astrophysics.
However, if we do indeed receive GWs from such a binary source, what information can we
possibly extract from it about the internal structure of neutron stars?
In this section we consider this question in perspective of measuring the tidal 
deformability parameter of neutron stars purely from gravitational waves they emits 
while orbiting spinning black holes.
We find that when the BH is sufficiently spinning in the prograde sense,
with $\chibh\gtrsim +0.7$, even at a relatively possible signal strengths ($\rho=20$),
we can begin to constraint $\lambdans$ at the level of $\pm 75\%$ errors with a $90\%$
confidence level. If the signal is stronger, say $\rho=30$, we can contraint
$\lambdans$ better with only $\pm 50\%$ errors, for binaries with large BH spins.
This trend continues as we vary the SNR from $\rho=30-50$ upwards. In other words,
with a single but moderately loud NSBH signal observation, Advanced LIGO will begin
to put constraints on NS matter properties. These constraints can consequently
be used to assess the likelihood of different equations of state for nuclear
matter, and possibly narrow the range they span.

  `

We use the same set of systems as in the previous section, with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, $\lambdans=\{500, 800, 1000, 1500, 2000\}$ and SNRs
$\rho=\{20,30,50,70\}$. For each unique combination of these parameters, we obtain a
signal waveform using the tidal model of Lackey et al~\cite{Lackey:2013axa}.
We use the same waveform model as input to our Bayesian parameter estimation
method in order to obtain estimates for different binary parameters alongwith 
their full probability distributions, including for the combined NS tidal
deformability parameter $\lambdans$.

The precision with which $\lambdans$ can be measured is given by the width of 
its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.

In Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we take the case of
faily deformable neutron stars, with actual $\lambdans = 1500$ (left panel) and
$2000$ (right panel). In each case, we calculate the lowest SNR value $\rho_c$ 
in order
for the signal to have enough information that we can constrain $\dlambda$ below
$100\%$. Note that this is a modest requirement on constrainability of $\lambdans$.
In each panel of the figure, we show $\rho_c$ as a function of BH mass and spin.
% 
We immediately note that to fulfil this modest aim with a single observation, 
we need an SNR of $30$ or higher in each detector. Given we have that, we can 
bring $\dlambda$ below $100\%$ for the lowest BH masses and higher spins. To
attain the same for other values of BH parameters, we will need a very loud
signal with $\rho\gtrsim 50$.

% 
% 
% \subsection{Constraining the tidal deformability of the neutron star}\label{s2:measuring_lambda}
%

% As before we focus on binaries with mass-ratios $q=\{[2,3,4,5\}$, with the NS mass
% fixed at $1.35M_\odot$. We allow BH spin to take the values 
% $\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000\}$. For each of these
% parameter combinations, we inject a tidal-corrected waveform using the 
% Lackey et al~\cite{} model , and use MCMC with ensemble sampling to construct the 
% posterior probability distribution for different binary parameters with (a) BHBH
% templates, and (b) NSBH templates. We will discuss the results of (b) here.
